---
title: "Simulating data with Directed Acyclic Graphs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating data with Directed Acyclic Graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r include=FALSE}
library(LST)
library(tibble)
```

Instructors often hear that students should work with "real" data, that is, data from actual observations or experiments. Such "real" data is used throughout *Lessons in Statistical Thinking* and serves several purposes: motivating students, creating verisimilitude, and exercising the statistical thinking skill of recognizing that one's findings from data may not reflect the reality that created the data.

But there is also an important use for "unreal," simulated data. The user-defined rules of the simulation stands for reality, providing a chance to see if the results of analyzing the data correspond with that reality. Another way to say this is that the use of simulated data allows one to look at **accuracy** and **bias**. As well, simulations provide a direct way to observe sampling variation, since a given simulation can be run multiple times.

In the `{LST}` package, data simulations are provided by a handful of purpose-written functions, particularly `datasim_make()` and `datasim_run()`. Simulation is always a two-step process: (1) create the rules by which the simulation works and (2) apply those rules using random-number generators to generate data. To illustrate the process, consider a very simple simulation in which a random `x` is generated, which is used to produce a `y` by a linear transformation of the `x`. The two simulated variables are then packaged as a data frame.

```{r}
Example1 <- datasim_make(
  x <- rnorm(n, mean = 0, sd = 2), 
  y <- -3 + 10 * x + rnorm(n, mean = 0, sd = 5)
)
Example1 |> sample(n=6,  seed = 101)
```

`Example1` is a simulation with two rules: `x` is a random normal variable (with the specified mean and standard deviation) and `y` is an arithmetic transformation of `x`.

For readers interested in details, we'll explain in @sec-why-datasim why the choice was taken to arrange things as this two-step process. For the present, however, we will continue without worrying about the design choices.

## Basics of `datasim_make()`

Each of the rules of a simulation is written in the syntax of an ordinary R statement. Within any one simulation objects, each rule must have a unique name. For example, `x <- rnorm(n, mean=0, sd=2)`, is a perfectly ordinary R statement, storing under the name `x` the rule `rnorm(n, mean=0, sd=2)`. 

`datasim_make()` is a **function**; the simulation rules are given as **arguments** to the function, that is, in parentheses and separated by commas. 

But there are three aspects of using `datasim_make()` that are entirely out of the ordinary. First, the name is given to each of the rules using the storage arrow (`<-`) as opposed to the usual `=` syntax for named arguments. Using `=` will generate an error, like this.

```{r error = TRUE}
Wrong_way <- datasim_make(x = rnorm(n, mean=0, sd=2)) # don't use =
```

Second, the quantity `n` does **not** need to be defined previously; it will be defined in the sampling phase.

Third, a rule may make use of a name (like `x`) only when that name has been used for a previous rule. For instance, the second rule in `Example1` is `y <- -3 + 10 * x` and uses the `x` defined in the first rule. More rules, each with its own unique name, can be added that make use of `x` and `y`.  

Once a set of rules has been collected using `datasim_make()`, the simulation can be run as many times as you like. 

```{r}
Example1 |> sample(n = 3)
Example1 |> sample(n = 2)
# and so on
```

## Pre-defined simulations

Although students or instructors may create their own simulations using `datasim_make()`, they will want to get started using some of the pre-defined simulations that come with the `{LST}` package. These are named with the prefix `sim_`, as in `sim_00`, `sim_01`, ..., `sim_12`, `sim_vaccine`, and so on.

There are two ways to see what are the rules of these predefined simulations.

1. Print out the rules, e.g.

```{r}
print(sim_04) # One of the simulations in {LST}
```

2. Draw a **graph** of the connections between rules, e.g. in `dag_04` rule `d` receives inputs from `a`, `b`, and `c`, but none of those other rules are connected to one another.

```{r label='DAGsMFIYTT'}
dag_draw(sim_04)
```

The term for the mathematical graphs corresponding to data simulation rules is "directed acyclic graph" (DAG, for short).

## Introducing causality

The rules of a data simulation form a network; some variables depend on others. In mathematics, networks are often represented by **graphs**, that is, systems with nodes and edges connecting the nodes. In data simulations, each rule specifies the inputs (if any) on the right-hand side of `<-` and how those inputs specify the value of the variable named on the left-hand side of `<-`. In other words, data simulations correspond to the mathematical idea of a **directed graph**.

But not all directed graphs are admissible. For instance, the following statement attempts to make a simulation with two variables, each of which sets the values of the other.

```{r error = TRUE}
cycle <- datasim_make(x <- y, y <- -x)
```

The mathematically oriented reader may be tempted to interpret the above as a set of *differential equation*, for instance the similar looking  $$\frac{dx}{dt} = y, \ \frac{dy}{dt} = -x$$. The differential equation is a legitimate mathematical object and has solutions, that is $x(t)$ and $y(t)$. And in the differential equation, $y$ clearly influences $x$ and *vice versa*. Or, to be precise, the *present$ value of $x$ influences the *future values* of $y$. (For R functions pertaining to differential equations, see the `{mosaicCalc}` package, which is not related to `{LST}` other than by authorship and affiliation with [Project MOSAIC]{https://www.mosaic-web.org/}.)

In a data simulation, in contrast, the rows in the output data are all independent realizations of the simulation rules. For instance, in `sim_04` (graphed above), for each row a value of `a` is generated autonomously ("exogenously"), as are the values of `b` and `c`. The `d` variable for that row is the arithmetic sum `a + b + c` with a bit of exogenous noise added as well.

The problem with the proposed simulation `cycle` is that to find `x` for a row, you must first know `y`, and to find `y` for a row, you must already know `x`. In order to avoid the impossible situation of a cycle in a data simulation, each successive rule can depend only on the rules that come before it. Such a network is called **acyclic**: without cycles. Consequently, data simulations correspond to Directed Acyclic Graphs (DAGs).

As you may know, DAGs play an important role in statistical **causal inference**. Consequently, DAGs are an important theme in *Lessons in Statistical Thinking*.

Data simulations can be useful in teaching causal concepts. To illustrate, consider the `sim_vaccine` simulation provided by the `{LST}` package. The simulation is a cartoon about how flu vaccinations might reduce mortality from the flu. We can look at `sim_vaccine` in the usual ways:

```{r label='DAGswTgvGw'}
print(sim_vaccine)
dag_draw(sim_vaccine)
```

To judge from the graph, which shows no arrows, there are no connections among the four variables involved in the simulation. This is a good starting point for a classroom discussion of how the four variables might be connected in the real world. Different people have different opinions about this, for example, "vaccine deniers" might insist that vaccination causes `health` to deteriorate. A person more in touch with medical science would see that vaccination effects the chances of getting the flu, and depending on the person's health a person with the flu is more likely to die.

We can run the simulation and model the resulting data to see if the data show any connections among the four variables. For instance, it's easy to simulate a study with 10,000 participants whose baseline health is measured and then are observed: do they get vaccinated, do they get the flu, do they die?

```{r}
sim_data <- sim_vaccine |> sample(n = 10000)
```

Modeling the data helps us address questions of interest. For instance, is vaccination associated with reduced odds of getting the flu?

```{r}
sim_data |> model_train(flu ~ vaccinated) |> conf_interval()
```

The negative coefficient on `vaccinated` indicates that vaccination (in the simulation) is associated with lower odds of getting the flu. Similarly, getting the flu is associated with dieing:

```{r}
sim_data |> model_train(zero_one(died, one="yes") ~ flu) |> conf_interval()
```

The positive coefficient on `flu` demonstrates the association: getting flu means larger odds of dieing.

You might wonder how associations are formed among the variables in the simulation when the graph has no edges connecting them. You can write your own simulation to verify that unconnected variables almost always produce a confidence interval that includes zero. Only about 5% of such simulation runs will result in a confidence interval not including zero. These runs lead to Type I errors.

The reason for the associations seen in the `sim_vaccine` data is that there are in fact connections, but they involve variables in the simulation that aren't being reported in the data, so called "lurking" variables. Unlike the real world, in the simulation we can look at such variables

```{r label='DAGsJsATe2'}
print(sim_vaccine, report_hidden = TRUE)
dag_draw(sim_vaccine, report_hidden = TRUE)
```

The variables `health`, `vaccinated`, `flu`, and `died` are merely readouts of internal variables `.h`, `.v`, `.f` and `.s`. Dots starting a variable name are what keeps those variables hidden, until revealed by the `report_hidden = TRUE` argument. 

The hidden variables are a nest of connections. For example, better health (`.h`) makes vaccination (`.v`) more likely.

To generate simulated data that includes the hidden variables, use the `report_hidden=TRUE` argument to `sample()`.

In a proper clinical trial, vaccination would be assigned at random, or maybe even blocked with respect to health, age, .... As it stands, `sim_vaccine` is an observational study. To create a randomized clinical trial---that is, an experiment---we would need to intervene to set `.v`. This can be done with the `datasim_intervene()` function, which allows us to specify new rules and modify existing ones. Like this:

```{r}
Randomized_trial_sim <- datasim_intervene(sim_vaccine, .v <- c(-5, 5))
dag_draw(Randomized_trial_sim, report_hidden = TRUE)
```

Note how the graph of `RCT`differs from the graph of that of `sim_vaccine`: there is now no direct connection between the health variable and vaccination. Generating data from the randomized clinical trial and modeling the connections between the flu variable (`.f`) and the vaccinated variable (`.v`) shows a much smaller connection between vaccination and the flu than was seen with the observational data generated by `sim_vaccine`.

```{r}
RTdata <- Randomized_trial_sim |> sample(n=10000, report_hidden = TRUE)
RTdata |> model_train(.f ~ .v + .h) |> conf_interval()
```

Of course, `sim_vaccine` is just a simulation and may not accord with how things work in the real world. But if the real world were like `sim_vaccine`, we would be able to untangle the confounded connections by running the experiment. Such untangling can also be done by the appropriate choice of covariates in a model.

## Why `datasim_make()`? {#sec-why-datasim}

Many textbooks that use R simulations arrange the code as a sequence of R statements. For instance, the simulation `Example1` shown at the start of this document could be written in three statements.

```{r}
n <- 100
x <- rnorm(n, mean = 0, sd = 2) 
y <- -3 + 10 * x + rnorm(n, mean = 0, sd = 5)
```

This approach is viable in terms of the numbers generated for `x` and `y`, but it has several shortcomings.

- The variables have not been assembled in a data frame, so graphing, summarizing, and modeling of the variables has to be done accordingly.
- There is no name for the simulation itself, so referring to a simulation or comparing simulations with different rules is awkward.
- Repeated runs of the simulation are tedious and book-keeping has to be done by hand.
- Analyzing the rules to draw a DAG or to check for cycles is not easy.

`datasim_make()` and `sample()` avoids such problems with a very small minimum of extra code, as seen in `Example1`.

## Functions for simulation rules

Many of the rules in simulation can be expressed in mainstream R code. For example, random numbers can be generated with `rnorm()`, `rexp()`, `rbinom()`, and the many other such functions provided by the `{stats}` package or other packages. Similarly, arithmetic and other mathematical operations can be written in the ordinary way.

It is convenient, however, to use special functions provided by `{LST}` for writing simulations. These functions simplify any common situations in simulations.

- `categorical()` generates specified categorical levels at random. We'll give examples using `n = 10`

```{r}
n <- 10
color <- categorical(n, "red", "blue", "green")
color
```      
By default, each level is equally likely to occur. You can specify probabilities directly using an argument notation:

```{r}
color <- categorical(n, red = .1, blue = .2, green = .7)
color
```

- `bernoulli()` generates zero-one variables or two-level categorical variables. The simple default is to make 0 and 1 with equal probability.

```{r}
flip <- bernoulli(n)
flip
```

You can provide labels for zero and one, producing a categorical variable:

```{r}
flip <- bernoulli(n, labels = c("bad", "good"))
flip
```

`bernoulli()` can also generate outcomes that depend on other variables, using those variables as either log-odds inputs or probability [0,1] inputs. See the `sim_vaccine` presented above for examples.

- `cat2value()` Sometimes you want to use a categorical variable as an input and translate it to numerical values as an output or, for instance, to specify log-odds in `bernoulli()`.

```{r}
vals <- cat2value(flip, good = 2, bad = -1 )
vals
result <- bernoulli(logodds = vals, labels = c("fail", "succeed"))
result
```

If you don't need `vals` elsewhere in your simulation, you can combine the two statements into one:

```{r}
result <- bernoulli(logodds = cat2value(flip, good=2, bad = -1))
result
```

- `block_by()` assigns categorical levels randomly, but arranged in blocks defined by some other variable. 

```{r}
treatment <- block_by(color, levels = c("glue", "nails", "screws"))
treatment
```

Using these same rules in `datasim_make()` facilitates running the simulation multiple times and packing each simulation's data as a data frame. To illustrate, it's hard to see from the `treatment` variable alone that it has been blocked by `color`. But once the data are in a data frame, we can use wrangling or modeling to reveal such patterns. (Don't forget the commas between the rules! Each rule is an argument to `datasim_make()`.)

```{r}
all_together <- datasim_make(
  color <- categorical(n, red = .1, blue = .2, green = .7),
  flip <- bernoulli(n, labels = c("bad", "good")),
  result <- bernoulli(logodds = cat2value(flip, good=2, bad = -1)),
  treatment <- block_by(color, levels = c("glue", "nails", "screws"))
)
all_together |> sample(n = 20) |> arrange(color)
```

In the above data frame, you can see that all the treatments have been assigned as evenly as possible for each of the colors.

- `mix_with()` will mix noise with a signal so that the output has the specified R^2^ and variance. 

```{r}
Mix_them <- datasim_make(
  x <- rnorm(n),
  y <- mix_with(x, R2 = 0.71, var = 4.35)
)
Dat <- Mix_them |> sample(n = 100)
Dat |> model_train(y ~ x) |> R2()
Dat |> summarize(var(y))
```
The noise being mixed with the signal may incidentally be correlated with the signal. This leads to R^2^ varying around the specified value. You can avoid this variation (not that you necessarily want to) by providing the optional argument `exact = TRUE`.

## Doubly stochastic processes

There are many situations where the conditions for a random outcome are themselves set at random. The `bernoulli()` process provides an illustration. Suppose that an outcome ill-vs-healthy is to be determined at random, but the patient's general health condition sets the probability of falling ill. For instance:

```{r}
Sick_sim <- datasim_make(
  health <- rnorm(n, sd=2), 
  sick   <- bernoulli(logodds = health, labels = c("ill", "OK"))
)
Sim_dat <- Sick_sim |> sample(n = 1000)
Sim_dat |> point_plot(sick ~ health)
```

A hallmark of such processes is that the `n` argument isn't needed; the number of random values produced is determined by the length of an input variable, in this case `health`.

`bernoulli()` was written specially to achieve this behavior. It's helpful to have a mechanism that will enable any doubly stochastic behavior. For instance, suppose we want to generate `k` a random positive integer and based on `k`, another variable `y` which will be the sum of `k` normal processes. It's tempting to try to write this as follows:

```{r}
# This won't work!
sum_sim <- datasim_make(
  k <- sample(c(1, 4, 16, 32, 64, 128), n = n, replace = TRUE),
  y <- sum(rnorm(k))
)
sum_sim |> sample(n = 5)
```

This hasn't performed as might be expected. All the supposedly random `y` values are the same. The reason can be seen from the documentation for `rnorm()`. When the `n` argument to `rnorm()` is a vector---and `k` will be a vector of length n = 5 when the sample is generated---the length of the vector determines the number of random numbers generated. Consequently, `sum(rnorm(k))` is a simple sum of n random numbers, rather than a n sums of a varying number of random numbers.

The `each()` function sets things up so that a separate calculation is done for each of the n rows in the sample. Each such calculation draws on the values of the variables for that particular row. To illustrate:

```{r}
# This will do what we want!
sum_sim <- datasim_make(
  k <- sample(c(1, 4, 16, 32, 64, 128), n = n, replace = TRUE),
  y <- each(sum(rnorm(k)))
)
sum_sim |> sample(n = 5)
```

We would need a bigger sample to see the `k` dependence of `y`:

```{r}
sum_sim |> sample(n = 10000) |>
  point_plot(y ~ k, point_ink = 0.1)
```
